How the Practice of TDD Influences \\Class Design in Object-Oriented Systems: \\Patterns of Unit Tests Feedback

### 2012-11-27 sbes2012-english/paper.tex
slides sbes; correcoes do ingles

\begin{abstract} 
	Despite that Test-Driven Development (TDD) appears to be a software testing practice, 
	many developers affirm that the practice influence on class design. This study aimed
	to better understand the effects of TDD and how the practice influences developers
	during class design on object-oriented systems. 
	We conducted an essentially qualitative exploratory study in which participants
	were invited to implement some exercises using TDD and, based on the data gathered,
	we gathered details of how the practice influenced design decisions from the
	participants using interviews.
	We noticed that the practice of TDD drives
	developers during class design by means of constant feedback about its quality. This study
	also named and catalogued feedback patterns perceived by the developers.
\end{abstract}

@@

The adoption of TDD by the industry has been growing. In a 2010 survey to discover
which practices were used by agile teams, Scott Ambler showerd that 53\% of all teams
that replied the survey adopted TDD as a way to validate the work done
\cite{wambler-survey-agile}. Similar numbers can be found on the annual surveys from
Version One that, in the 2012 version \cite{versionone-2012}, showed that
40\% of the respondent teams have been using the practice.

When using TDD, a developer only writes code that is covered by a test. Because
of that, it is common to relate the practice of TDD with software testing. However,
a common speech among TDD practitioners is its effects on the internal code quality.
Many authors, well-known by industry and academia, such as Kent Beck \cite{TDDByExample}, 
Robert Martin \cite{agile-ppp}, Steve Freeman \cite{GOOS}, and Dave Astels \cite{astels-tdd}, 
state (without scientific evidence) that the practice of TDD promotes a significative
improvement on class design, helping the developer to create more cohesive less coupled
classes.

On the other hand, the way in which the practice of TDD guides the developer
through class design is not clear. We observed that in our qualitative study
with TDD practitioners, inside of a brazilian agile event. We interview
ten participants about TDD effects \cite{aniche-wbma} and, for our surprise,
none of them could satisfactory explain how the practice guides them to better
designs.
Siniaalto and Abrahamsson \cite{alarming-results} also share this opinion. They also
noticed that the effects of TDD can not be as automatic as the expected for most people.
The related work, discussed in Section \ref{cap:trabalhos-relacionados},
only evaluates if the practice of TDD makes difference in the produced code. Only a few
of them contain a qualitative study that details how the practice does that difference.
With this information, developers would know how to use the practice of TDD in order to 
better create class design.

@@

Based on the fact that the software development process involves many
human factors and is totally sensitive to the context that it belongs,
we conduct an essencially qualitative exploratory study in which professionals
from industry were invited to implemnet some prepared exercises using TDD and,
based on the gathered data, we investigate the details of how the practice influenced
the participants' class design decisions through interviews.

@@

Test-Driven Development, popularized by Kent Beck through his book, \textit{TDD: By Example} (2011)
\cite{TDDByExample}, is one of the practices in which the focus is to give feedback. TDD has
a big importance during the software development cycle because, as suggested by the agile
practices, the class design of a system should emerge as software grows. And, in order
to quickly respond to this evolution, a constant feedback about internal and external code
quality is needed.

TDD is a software development practice that is based on a repetition of a small
cycle. First, developers should write a failing test. Then, they make it pass,
by implementing the desired functionality. After that, they refactor the code
to remove code or data duplication that was generated by the process. Besides that,
simplicity is intrisec to the process: the practitioner tries to write the simplest
test that fails and tries to write the simplest implementation that makes the
test pass. This cycle is also known as "Red-Green-Refactor", as it remembers the
colors that a developer usually sees when doing TDD: the red means that a test
is failing, and green means that the test passes succesfully.

It is common to relate TDD to software testing practices. Although the creation
of tests is intrisec to the process, it is also said that TDD helps developers
to create more flexible, cohesive, and less coupled classes. The tests are the
tool used by developers to evaluate the class design that is being developed.
Because of that, many developers refers to TDD 
as \textit{Test-Driven Design} \cite{tdd-taxonomy}.

Authors such as Kent Beck \cite{aim-fire}, Dave Astels \cite{astels-tdd}, and
Robert Martin \cite{bob-martin} state that TDD is, in fact, a class design
technique \cite{tdd-taxonomy} \cite{aim-fire}.
In their opinion, the change in the tradicional development cycle, although very simple,
aggregates many other benefits to the produced code: more simplicity, less coupling, and
a high cohesion. Ward Cunningham, one of the Extreme Programming pioneers, summarizes
the discussion in one statement: \textit{"Test-First programming is not a testing technique"} 
\cite{aim-fire}.
According to Janzen, a more clear definition is the one that says that TDD is the 
art of producing automated tests to production code, and use this process to guide
the class design and programming \cite{agilealliance-tdd} \cite{tdd-taxonomy}.

It is a known fact that class design tend to loose quality during the software evolution.
However, it is hard to discuss quality in terms of class design. In this study context,
we used the SOLID principles, detailed in Martin's work \cite{bob-martin}, which are
mentioned in the section below.

@@

It is said that a class design is \textit{rotting} when it becomes hard
to evolve, the code reuse becomes more complicated than replicate the piece
of code, or the cost of doing any changes in the calss design becomes high.
Martin \cite{bob-martin} enumerated a few symptoms of a rotten class design,
also known as \textit{class design smells}. They are similar to \textit{code smells},
except that they belong to a higher level: they are present in the general structure
of the software instead of being located in just a single piece of code.

These symptoms can be measured subjectively and, sometimes, objectively. Usually
they are caused by violations of one or more class design principles.
In this study, we make use of the symptoms that were enumarated by him: 
rigidity, fragility, imobility, viscosity, opacity, needless complexity and
needless repetition. Besides, we also refer to the principles known as SOLID:
Single Responsibility Principle (SRP), Open-Closed Principle (OCP),
Liskov Substitutive Principle (LSP), Interface Segregration Principle (ISP), and
Dependency Inversion Principle (DIP) \cite{bob-martin}.

@@

Many empirical studies have been done in order to evaluate the effects of TDD.
In most of them, the effects of the practice in the class design is not taken
into account, and only the effect on external quality are measured. In addition,
in contrary to what this study proposes, many of them chosed
to had a bigger control in the experiment. They also have done it inside of the academic
environment, with students that belong to the most different computer courses.

@@

Another study by Janzen \cite{janzen-saiedian} with three different group
of students (each one of them using a different approach: TDD, test last, no tests)
showed that the code produced by the TDD team made better use of object-oriented
concepts, and responsibilities were better splitted into different classes,
while other teams produced a more procedural code. Also, tested classes were
104\% less coupled than non-tested classes, and methods were 43\%, on average, less complex
than the non-tested ones.

Dogsa and Batic \cite{dogsa-batic} also found an improvement on class design when
using TDD. But, according to the authors, the improvement is a consequence of the simplicity
TDD aggregates to the process. They also affirmed that the test suites created during
the practice favors the constant code refactoring.

Li \cite{angela-li} proposed a qualitative study to understand the efficacy of TDD.
In a case study, she collected the perceptions of TDD practitioners about the
benefits of the practices. To achieve that, she interviewed five developers
from software companies in Auckland, Australia. The results of the interview
were analyzed and discussed in terms of code quality, software quality, and
programmer's productivity. Regarding code quality, Li concluded that TDD
guides developers to simpler and better designed classes. In addition, the main
factors that contribute to these benefits are the confidence to refactor and
modify code, a higher code coverage, a deeper understanding of the requirements,
the ease of compreehending code, a reduced error scope, and the elevated
satisfaction of the developers.

The TDD practitioner usually makes uses of other agile practices, such as
pair programming. It difficults the evaluation of the practice.
Madeyski \cite{madeyski-package-dependencies} observed the results
among groups that practiced TDD, groups that practiced pair programming,
and the combination between them, and he was not able to show a significant
difference between teams that use TDD and teams that use pair programming
in terms of class package dependency management. However, when combining
the results, he found that TDD can help managing dependencies in class level.
According to him, developers should use TDD, but they should be aware of 
problems in class design.

@@

As presented, only a few studies evaluated the effects of TDD on class design.
And, when they do, they only discuss what the effects are and not exactly
\textbf{how} TDD has influenced. Josefsson \cite{josefsson}, in his discussion
about the need of an architectural phase and the effects of TDD in this matter,
comes to the same conclusion. According to him, the studies about TDD that are
found in the current literature are very limited and, because of that, the
"known" effects of the practice in class design can not be explained. Based on
our literature review, we believe that this limitation still exists.

Most part of these studies also does not take into account the experience
of the developer that is practicing TDD. Usually they only discuss it
during threats to validity. Janzen \cite{janzen-phd}, in his PhD, perceived
that mature developers obtain more benefits from the practice, writing simpler
classes. Besides, mature developers tend to choose TDD more often than
less experienced developers.

@@

A qualitative research is a way to explore and understand the influence that
individuals or groups put on a social or human problem. The research process
involves emerging questions and procedures, data are usually gathered through
the point of view from the participants, with the analyzis done in an inductive
way, from a very specific to a general theme, and with the researcher interpreting
the meaning of these data. Data captured by qualitative studies are usually
represented by words and figures. The final report contains a flexible structure.
Researchers that are dedicated to this kind of research support the inductive style and
the importance of showing the complexity of the situation \cite{creswell}.

@@

To achieve our goal, we choosed to conduct an essencially 
exploratory qualitative study with developers in industry. Participants
were invited to implement a set of pre-prepared exercises. After that,
they were interview about how the practice influenced them in their
class design decisions. This section details the planning as well as
the data analysis process.

@@

The main goal of this study is \textbf{to understand the relationship between
the practice of TDD and class design decisions taken by programmers in object-oriented
systems}. To compreehend them, we answered the questions below:

@@

Participants from different software development companies in the brazilian
market were selected. The participants' profile is discussed in sub-section
\ref{sec:planejamento-participantes}. All of them were invited to implement
a few problems using Java during a defined timebox. Participants practiced
TDD in one problem and did not practice in the other. The problems
as well as the definition of each of them should be implemented using TDD were
randomized in order to reduce bias.

@@

To calculate these metrics, we implemented our own tool. The reason is that
most part of the existant tools make use of compiled code (instead of pure source
code). Our tool contains an automated test suite and an open source
\footnote{\url{http://www.metricminer.org.br}. 
Last access on July, the 6th, 2012.}.

@@

Developers from the brazilian software development market were invited to be part
of this research. The participants were evaluated according to their experience in TDD,
software development, Java, and unit tests. The only requirement was that the participant
should already knew how to write an unit test.

All participants filled out a survey before the start of the study. In this survey,
besides questioning their experience (quantitavely), it also contained open questions in which
the participant could tell about his experience in object-oriented systems, Java, and
TDD in a more detailed way.

@@

We can also affirm that half of the participants are still trying the practice. The others
are more experienced and have already consolidated the practice. This is positive, as it
was possible to capture informations from people with different levels of maturity.

@@

It was said to participants that the exercises simulate real world problems and that their code
would be supposely maintained by another team. Because of that, participants were asked to
implement the most elegant and flexible solution possible.

@@

		Exercise 1 & Rigidity, Needless Complexity & SRP, OCP \\
		Exercise 2 & Fragility, Viscosity, Imobility & SRP, DIP, OCP \\
		Exercise 3 & Rigidity, Fragility & SRP\\
		Exercise 4 & Fragility, Viscosity, Imobility & OCP, SRP, DIP \\

@@

To triangulate the information of the qualitative analysis, we calculated
code metrics to check whether there was a difference in code produced
with the practice of TDD.
We analysed
264 production classes (831 methods, 2520 lines) and
73 test classes (225 methods, 1832 lines).

@@

In Table \ref{metricas-industria}, we showed the \textit{p-values} found for
the difference between code produced with and without TDD. By the numbers,
we observed that there was no significant difference in cyclomatic complexity
and efferent coupling.
Lack of cohesion of methods was different in two exercises (1 and 4). A difference
also appeared on the quantity of lines in a method (exercise 4) and quantity of
methods (exercise 1).

@@

The values presented before corroborate with many of the related work.
Apparently TDD does not influence the code to the point of changing
the metric values for coupling, cohesion, and simplicity. However, this is
incoerent to the industry common feeling of the positive effects of TDD on class design.
As expected, in this study we conducted a qualitative analysis to understand
how this influence goes from the point of view of the developers.

@@

According to them, TDD can help during class design process but, in order to achieve that,
the developer should have certain experience in software development. Most of participants
affirmed that their class designs were based on their experiences and past learnings.
In their opinion, the best option is to join the practice of TDD and experience:

@@

Some participants did not finish their implementation. That can influence que
quantitative analysis, as a design that would be complex at the end, looks simple
to an automated code metric.

%% ------------------------------------------------------------------------- %%
\subsubsection{Reserarcher's infuence}

@@

Desirability bias is the scientific term used to describe the tendency of some participants
to reply questions in a way that they will be accepted by the other members
of community \cite{crowne}.
Agile methods and TDD have a strong speech. The brazilian agile community is still
young and it is possible to empirically perceive that many developers just
repeat the speech without a great experience or knowledge of the subject.

@@

Although we contacted many different companies and software development groups,
que number of participants invited may not be enough to generalize the findings.

@@

It is expected, with this cataloged patterns, that developers pay more attention
to the potential feedbacks that TDD brings and, use them to improving their class
design. The teaching of TDD can also be be assisted by this study, considering
that teachers can use the patterns raised here and show to students
how the practice can, actually guide the developer to a better class design.

A possible future work is the developing of tools that automatically detects this patterns
and warn the developer about a possible code smell in design. Besides that,
the search for more patterns like the showed here can improve even more the feedback
provided by the practice.

@@

The practice, through the possibles feedbacks on class design, heavily
discussed in Section \ref{padroes-tdd}, can be a guide to the developer.
These feedback, when noted, make the developer realize problems in class
design beforehand, facilitating code refactoring (Q1 and Q2).

@@

The developer that uses TDD writes tests before the code. This make
the current unit test to be treated as a draft by the developer.
\textbf{Noticing carefully the code of the unit test, the developer can
realize issues in class design that he is writing. Issues like 
classes with too many responsibilities or with too many dependencies}.

>>>>

\begin{abstract} 
	Despite the fact that Test-Driven Development (TDD) appears to be a software testing practice, 
	many developers affirm that the practice influence on class design. This study aimed
	to better understand the effects of TDD and how the practice influences developers
	during class design on object-oriented systems. 
	We conducted an essentially qualitative exploratory study in which participants
	were invited to implement some exercises using TDD and, based on the data gathered,
	we raised details of how the practice influenced design decisions from the
	participants using interviews.
	We noticed that the practice of TDD drives
	developers during class design by means of constant feedback about its quality. This study
	also named and catalogued feedback patterns perceived by the developers.
\end{abstract}

@@

The adoption of TDD by the industry has been growing. In a 2010 survey to discover
which practices were used by agile teams, Scott Ambler showed that 53\% of all teams
that replied the survey adopted TDD as a way to validate the work done
\cite{wambler-survey-agile}. Similar numbers can be found on the annual surveys from
Version One that, in the 2012 version \cite{versionone-2012}, showed that
40\% of the respondent teams have been using the practice.

When using TDD, a developer only writes code that is covered by a test. Due
to that, it is common to relate the practice of TDD with software testing. However,
a common speech among TDD practitioners is its effects on the internal code quality.
Many authors, well-known by industry and academia, such as Kent Beck \cite{TDDByExample}, 
Robert Martin \cite{agile-ppp}, Steve Freeman \cite{GOOS}, and Dave Astels \cite{astels-tdd}, 
state (without scientific evidence) that the practice of TDD promotes a significative
improvement on class design, helping the developer to create more cohesive less coupled
classes.

On the other hand, the way in which the practice of TDD guides the developer
through class design is not clear. We observed that in our qualitative study
with TDD practitioners, inside of a Brazilian agile event. We interview
ten participants about TDD effects \cite{aniche-wbma} and, for our surprise,
none of them could satisfactory explain how the practice guides them to better
designs.
Siniaalto and Abrahamsson \cite{alarming-results} also share this opinion. They also
noticed that the effects of TDD can not be as automatic as the expected for most people.
The related work, discussed in Section \ref{cap:trabalhos-relacionados},
only evaluates if the practice of TDD makes difference in the produced code. Only a few
of them contain a qualitative study that details how the practice does that difference.
With this information, developers would know how to use the practice of TDD in order to 
better create class design.

@@

Based on the fact that the software development process involves many
human factors and is totally sensitive to the context that it belongs,
we conduct an essentially qualitative exploratory study in which professionals
from industry were invited to implement some prepared exercises using TDD and,
based on the gathered data, we investigate the details of how the practice influenced
the participants' class design decisions through interviews.

@@

Test-Driven Development, popularized by Kent Beck through his book, \textit{TDD: By Example} (2011)
\cite{TDDByExample}, is one of the practices in which the focus is to give feedback. TDD has
a big importance during the software development cycle because, as suggested by the agile
practices, the class design of a system should emerge as software grows. And, in order
to quickly respond to this evolution, a constant feedback about internal and external code
quality is required.

TDD is a software development practice that is based on a repetition of a small
cycle. First, developers should write a failing test. Then, they make it pass,
by implementing the desired functionality. After that, they refactor the code
to remove code or data duplication that was generated by the process. Besides that,
simplicity is intrinsic to the process: the practitioner tries to write the simplest
test that fails and tries to write the simplest implementation that makes the
test pass. This cycle is also known as "Red-Green-Refactor", as it remembers the
colors that a developer usually sees when doing TDD: the red means that a test
is failing, and green means that the test passes successfully.

It is common to relate TDD to software testing practices. Although the creation
of tests is intrinsic to the process, it is also said that TDD helps developers
to create more flexible, cohesive, and less coupled classes. The tests are the
tool used by developers to evaluate the class design that is being developed.
Therefore, many developers refers to TDD 
as \textit{Test-Driven Design} \cite{tdd-taxonomy}.

Authors such as Kent Beck \cite{aim-fire}, Dave Astels \cite{astels-tdd}, and
Robert Martin \cite{bob-martin} state that TDD is, in fact, a class design
technique \cite{tdd-taxonomy} \cite{aim-fire}.
In their opinion, the change in the traditional development cycle, although very simple,
aggregates many other benefits to the produced code: more simplicity, less coupling, and
a high cohesion. Ward Cunningham, one of the Extreme Programming pioneers, summarizes
the discussion in one statement: \textit{"Test-First programming is not a testing technique"} 
\cite{aim-fire}.
According to Janzen, a more clear definition is the one that says that TDD is the 
art of producing automated tests to production code, and use this process to guide
the class design and programming \cite{agilealliance-tdd} \cite{tdd-taxonomy}.

It is a known fact that class design tend to lose quality during the software evolution.
However, it is hard to discuss quality in terms of class design. In this study context,
we used the SOLID principles, detailed in Martin's work \cite{bob-martin}, which are
mentioned in the section below.

@@

It is said that a class design is \textit{rotting} when it becomes hard
to evolve, the code reuse becomes more complicated than replicate the piece
of code, or the cost of doing any changes in the class design becomes high.
Martin \cite{bob-martin} enumerated a few symptoms of a rotten class design,
also known as \textit{class design smells}. They are similar to \textit{code smells},
except that they belong to a higher level: they are present in the general structure
of the software instead of being located in just a single piece of code.

These symptoms can be measured subjectively and, sometimes, objectively. Usually
they are caused by violations of one or more class design principles.
In this study, we make use of the symptoms that were enumerated by him: 
rigidity, fragility, immobility, viscosity, opacity, needless complexity and
needless repetition. Besides, we also refer to the principles known as SOLID:
Single Responsibility Principle (SRP), Open-Closed Principle (OCP),
Liskov Substitutive Principle (LSP), Interface Segregation Principle (ISP), and
Dependency Inversion Principle (DIP) \cite{bob-martin}.

@@

Many empirical studies have been done in order to evaluate the effects of TDD.
In most of them, the effects of the practice in the class design is not taken
into account, and only the effect on external quality are measured. In addition,
in contrary to what this study proposes, many of them chosed
to have a bigger control in the experiment. They also have done it inside of the academic
environment, with students that belong to the most different computer courses.

@@

Another study by Janzen \cite{janzen-saiedian} with three different groups
of students (each one of them using a different approach: TDD, test last, no tests)
showed that the code produced by the TDD team made better use of object-oriented
concepts, and responsibilities were better divided into different classes,
while other teams produced a more procedural code. Also, tested classes were
104\% less coupled than non-tested classes, and methods were 43\%, on average, less complex
than the non-tested ones.

Dogsa and Batic \cite{dogsa-batic} also found an improvement on class design when
using TDD. However, according to the authors, the improvement is a consequence of the simplicity
TDD aggregates to the process. They also affirmed that the test suites created during
the practice favors the constant code refactoring.

Li \cite{angela-li} proposed a qualitative study to understand the efficacy of TDD.
In a case study, she collected the perceptions of TDD practitioners about the
benefits of the practices. To achieve that, she interviewed five developers
from software companies in Auckland, Australia. The results of the interview
were analyzed and discussed in terms of code quality, software quality, and
programmer's productivity. Regarding code quality, Li concluded that TDD
guides developers to simpler and better designed classes. In addition, the main
factors that contribute to these benefits are the confidence to refactor and
modify code, a higher code coverage, a deeper understanding of the requirements,
the ease of comprehending code, a reduced error scope, and the elevated
satisfaction of the developers.

The TDD practitioner usually makes uses of other agile practices, such as
pair programming. It makes the evaluation of the practice more difficult.
Madeyski \cite{madeyski-package-dependencies} observed the results
among groups that practiced TDD, groups that practiced pair programming,
and the combination between them, and he was not able to show a significant
difference between teams that use TDD and teams that use pair programming
in terms of class package dependency management. However, when combining
the results, he found that TDD can help managing dependencies in class level.
According to him, developers should use TDD, but they should be aware of 
problems in class design.

@@

As presented, only a few studies evaluated the effects of TDD on class design.
And, when they do, they only discuss what the effects are and not exactly
\textbf{how} TDD has influenced them. Josefsson \cite{josefsson}, in his discussion
about the need of an architectural phase and the effects of TDD in this matter,
comes to the same conclusion. According to him, the studies about TDD that are
found in the current literature are very limited and, because of that, the
"known" effects of the practice in class design can not be explained. Based on
our literature review, we believe that this limitation still exists.

Most part of these studies also does not take into account the experience
of the developer who is practicing TDD. Usually they only discuss it
during threats to validity. Janzen \cite{janzen-phd}, in his PhD, perceived
that mature developers obtain more benefits from the practice, writing simpler
classes. Besides, mature developers tend to choose TDD more often than
less experienced developers.

@@

A qualitative research is a way to explore and understand the influence that
individuals or groups put on a social or human problem. The research process
involves emerging questions and procedures, data are usually gathered through
the point of view from the participants, with the analysis done in an inductive
way, from a very specific to a general theme, and with the researcher interpreting
the meaning of these data. Data captured by qualitative studies are usually
represented by words and figures. The final report contains a flexible structure.
Researchers that are dedicated to this kind of research support the inductive style and
the importance of showing the complexity of the situation \cite{creswell}.

@@

To achieve our goal, we chose to conduct an essentially 
exploratory qualitative study with developers in industry. Participants
were invited to implement a set of pre-prepared exercises. After that,
they were interview about how the practice influenced them in their
class design decisions. This section details the planning as well as
the data analysis process.

@@

The main goal of this study is \textbf{to understand the relationship between
the practice of TDD and class design decisions taken by programmers in object-oriented
systems}. To comprehend them, we answered the questions below:

@@

Participants from different software development companies in the Brazilian
market were selected. The participants' profile is discussed in sub-section
\ref{sec:planejamento-participantes}. All of them were invited to implement
a few problems using Java during a defined timebox. Participants practiced
TDD in one problem and did not practice in the other. The problems
as well as the definition of each of them should be implemented using TDD were
randomized in order to reduce bias.

@@

To calculate these metrics, we implemented our own tool. The reason is that
most part of the existent tools make use of compiled code (instead of pure source
code). Our tool contains an automated test suite and an open source
\footnote{\url{http://www.metricminer.org.br}. 
Last access on July, the 6th, 2012.}.

@@

Developers from the Brazilian software development market were invited to be part
of this research. The participants were evaluated according to their experience in TDD,
software development, Java, and unit tests. The only requirement was that the participant
should already knew how to write a unit test.

All participants filled out a survey before the start of the study. In this survey,
besides questioning their experience (quantitatively), it also contained open questions in which
the participant could tell about his experience in object-oriented systems, Java, and
TDD in a more detailed way.

@@

We can also affirm that half of the participants are still trying the practice. The others
are more experienced and have already consolidated the practice. This is positive, as it
was possible to capture information from people with different levels of maturity.

@@

It was said to participants that the exercises simulate real world problems and that their code
would be supposedly maintained by another team. Because of that, participants were asked to
implement the most elegant and flexible solution possible.

@@

		Exercise 1 & Rigidity, Needless Complexity & SRP, OCP \\
		Exercise 2 & Fragility, Viscosity, Immobility & SRP, DIP, OCP \\
		Exercise 3 & Rigidity, Fragility & SRP\\
		Exercise 4 & Fragility, Viscosity, Immobility & OCP, SRP, DIP \\

@@

To triangulate the information of the qualitative analysis, we calculated
code metrics to check whether there was a difference in code produced
with the practice of TDD.
We analyzed
264 production classes (831 methods, 2520 lines) and
73 test classes (225 methods, 1832 lines).

@@

In Table \ref{metricas-industria}, we showed the \textit{p-values} found for
the difference between codes produced with and without TDD. By the numbers,
we observed that there was no significant difference in cyclomatic complexity
and efferent coupling.
Lack of cohesion of methods was different in two exercises (1 and 4). A difference
also appeared on the quantity of lines in a method (exercise 4) and quantity of
methods (exercise 1).

@@

The values presented before corroborate with many of the related work.
Apparently TDD does not influence the code to the point of changing
the metric values for coupling, cohesion, and simplicity. However, this is
incoherent to the industry common feeling of the positive effects of TDD on class design.
As expected, in this study we conducted a qualitative analysis to understand
how this influence goes from the point of view of the developers.

@@

According to them, TDD can help during class design process but, in order to achieve that,
the developer should have certain experience in software development. Most of participants
affirmed that their class designs were based on their experiences and past learning process.
In their opinion, the best option is to join the practice of TDD and experience:

@@

Some participants did not finish their implementation. That can influence the
quantitative analysis, as a design that would be complex at the end, looks simple
to an automated code metric.

%% ------------------------------------------------------------------------- %%
\subsubsection{Reserarcher's influence}

@@

Desirability bias is the scientific term used to describe the tendency of some participants
to reply questions in a way that they will be accepted by the other members
of community \cite{crowne}.
Agile methods and TDD have a strong speech. The Brazilian agile community is still
young and it is possible to empirically perceive that many developers just
repeat the speech without a great experience or knowledge of the subject.

@@

Although we contacted many different companies and software development groups,
the number of participants invited may not be enough to generalize the findings.

@@

It is expected, with this cataloged patterns, that developers pay more attention
to the potential feedbacks that TDD brings and, use them to improving their class
design. The teaching of TDD can also be assisted by this study, considering
that teachers can use the patterns raised here and show to students
how the practice can, actually guide the developer to a better class design.

A possible future work is the developing of tools that automatically detects this patterns
and warn the developer about a possible code smell in design. Besides that,
the search for more patterns like the ones shown here can improve even more the feedback
provided by the practice.

@@

The practice, through the possible feedbacks on class design, heavily
discussed in Section \ref{padroes-tdd}, can be a guide to the developer.
These feedbacks, when noted, make the developer realize problems in class
design beforehand, facilitating code refactoring (Q1 and Q2).

@@

The developer that uses TDD writes tests before the code. This makes
the current unit test to be treated as a draft by the developer.
\textbf{Noticing carefully the code of the unit test, the developer can
realize issues in class design that he is writing. Issues like 
classes with too many responsibilities or with too many dependencies}.
